---
title: "Data"
---

## Packages

In the short course, we'll actively use the `tidyverse` collection of packages for data manipulation; the `flexdashboard` package for creating dashboards; and the `shiny` package for interactivity. Code to install the packages follows:

```{r, eval=FALSE}
install.packages(c("tidyverse", "flexdashboard", "shiny"))
```

## Data

This shortcourse uses two main data resources. 

### FAS Litters

The `FAS_litters.csv` dataset is drawn from experiment in mice that explored the effect of in utero exposure to alcohol on development. The data contain information regarding the exposure group; a litter identifier; information regarding gestational weight gain and the gestational day on which the mother gave birth; and the postnatal day on which markers of development were observed. 

These data are primarily used to explore data import, manipulation, and visualization.

### NYC AirBNB

We will use data on AirBNB rentals in NYC in lectures creating dashboards and shiny apps. These data were accessed from [Inside Airbnb](http://insideairbnb.com/get-the-data.html) on September 2, 2017. 

The version of the data that we will use in this lecture can be found [here](./data/nyc_airbnb.zip), and then loaded using `readr::read_csv()`. You can also access these data through the `p8105.datasets` package. Install that package using:

```{r, eval=FALSE}
devtools::install_github("p8105/p8105.datasets")
```

then load the data with

```{r, eval = FALSE}
library(p8105.datasets)

data(nyc_airbnb)
```


**About [Inside Airbnb](http://insideairbnb.com/about.html)**

_Inside Airbnb is an independent, non-commercial set of tools and data that allows you to explore how Airbnb is really being used in cities around the world._

_By analyzing publicly available information about a city's Airbnb's listings, Inside Airbnb provides filters and key metrics so you can see how Airbnb is being used to compete with the residential housing market._


## Data description

Inside Airbnb provides some visualizations of the NYC Airbnb data [here](http://insideairbnb.com/new-york-city/), where you can see maps showing type of room, activity, availability, and listings per host for all NYC Airbnb listings.

The resulting R data file `nyc_airbnb` contains a single dataframe `df` with 40,753 rows of data on 17 variables:

- `id`: listing id
- `review_scores_location`: 0-5 stars converted into a 0-10 scale
- `name`: listing name
- `host_id`: host id 
- `host_name`: host name
- `neighbourhood_group`: NYC borough
- `neighbourhood`: NYC neighborhood
- `latitude`: listing latitude
- `longitude`: listing longitude
- `room_type`: type of listing (Entire home/apt, Private room, Shared room)
- `price`: listing price
- `minimum_nights`: required minimum nights stay
- `number_of_reviews`: total number of reviews
- `last_review`: date of last review
- `reviews per month`: average number of reviews per month
- `calculated_host_listings_count`: total number of listings for this host
- `availability_365`: number of days listing is available out of 365


